---
title: Lets make a map in R
author: ~
date: '2020-07-13'
slug: []
categories: []
tags: []
lastmod: '2020-11-18T20:57:33Z'
keywords: []
description: ''
comment: no
postMetaInFooter: no
hiddenFromHomePage: no
contentCopyright: no
reward: no
mathjax: yes
mathjaxEnableSingleDollar: no
mathjaxEnableAutoNumber: no
hideHeaderAndFooter: no
flowchartDiagrams:
  enable: no
  options: ''
sequenceDiagrams:
  enable: no
  options: ''
---



There's a lot of really great posts on mapping in R. [This one](https://medium.com/@traffordDataLab/lets-make-a-map-in-r-7bd1d9366098) from the [Trafford Data Lab](https://www.trafforddatalab.io/) is the one I used when I was starting out in R, and is the post I recommend to anyone interested in making maps with R.

Most posts use `ggplot2` to make maps. Here, we're going to use the `tmap` package. `tmap` makes **t**hematic **maps** - which is a fancy way of saying 'plotting data on a map'. I prefer `tmap` to `ggplot2` because tmap was built specifically for making maps. `ggplot2` is a general plotting library - while it can make maps, it's usually faster & easier to make them in `tmap`.

This post is a quick overview of using `tmap` (plus a couple of other libraries). We're going to:

* Make a choropleth map
* Add in a basemap using OpenStreetMap

If you want more info, have a look at the [tmap vignette](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html).

First thing's first, we're going to load the packages needed for mapping:

```{r warning=F, message=F}
library(dplyr)         # For working with data
library(sf)            # For working with shapefiles
library(tmap)          # For making maps
library(cartography)   # For downloading basemaps
library(raster)        # For plotting basemaps
library(fingertipsR)   # For health data
```

# Lets make a choropleth

We're going to make a choropleth of obesity prevalence across England. A choropleth is just a colourful map - the map is divided up into different areas, then those areas are coloured in based on some variable. Before we can make a choropleth we're going to need two things: a map, and some data. These are both dead easy to get. 

We'll start with the map. We need to get a *shapefile* (a file which tells the computer how to draw the map). In the UK the best place to go is the [ONS Geoportal](https://geoportal.statistics.gov.uk/), which contains all the shapefiles used by the ONS. We're going to grab the [ultra-generalised CCG boundaries](https://geoportal.statistics.gov.uk/datasets/clinical-commissioning-groups-april-2019-boundaries-en-buc). A CCG is a clinical commissioning group, they're responsible for health related things so any health data (like obesity prevalence) is usually displayed at CCG level. Ultra-generalised shapefiles are slightly less detailed than other shapefiles. This is fine for plots, and they're usually smaller files so are quite fast to work with.


Head over to the link and click on the 'APIs' button. Highlight the 'GeoJSON' one and read it in:

```{r warning=F, message=F}
shp_ccg = read_sf('https://bit.ly/3pHX0t1')
```

And use tmap to have a look at it. `tmap` is based on the layered grammar of graphics, same as `ggplot2`. This means the `tmap` syntax is very similar to `ggplot2`:

```{r warning=F, message=F}
tmap_mode('plot')      # Tell tmap to do a plot
tm_shape(shp_ccg) +    # Using this shape...
  tm_borders()         # ... plot the borders
```

We can get the data using the excellent `fingertipsR` package from Public Health England:

```{r}
df_prev = fingertips_data(IndicatorID = 92588, 
			  AreaTypeID = 165)
```

This returns the yearly obesity prevalence figures for each CCG. The prevalence is in the `Value` column. Lets grab the latest numbers:

```{r}
df_prev = df_prev %>%
  filter(Timeperiod == '2018/19')
```

Now all we need to do is join the data together and we're ready to plot. Looking at the names of `shp_ccg` and `df_prev`, we can join on the CCG code - called `CCG19CD` in `shp_ccg` and `AreaCode` in `df_prev`:

```{r warning=F, message=F}
shp_ccg = shp_ccg %>%
  left_join(df_prev, by = c('CCG19CD' = 'AreaCode'))
```

Plotting this is easy enough - we just need to tell `tmap` to plot polygons (rather than borders like we did above) & tell it how to colour those polygons in:

```{r warning=F, message=F}
tmap_mode('plot')
tm_shape(shp_ccg) +
  tm_polygons(col = 'Value')
```

And we're done! As you can see, `tmap` figures out some reasonable guesses at how many categories to use. You can change this using the `n = ` argument if you want to use a specific number of categories. You can also change how the categories are chosen by changing the `style = ` argument. A good default is to use `style = 'jenks'`, which tends to make a smallish number of classes & handles outliers well:

```{r warning=F, message=F}
tmap_mode('plot')
tm_shape(shp_ccg) +
  tm_polygons(col = 'Value', 
              style = 'jenks', 
              title = 'Obesity prevalence %',
              palette = 'viridis')

```

We've customised the plot using the `title` and `palette` arguments. There's loads of other arguments you can mess about with if you want to customise the map further. Type in `?tm_polygons` to read more about them.

# Lets add a basemap

Everything we've seen so far can be done in ggplot (go read the [Trafford Data Labs blog]((https://medium.com/@traffordDataLab/lets-make-a-map-in-r-7bd1d9366098)) to see how!). But there's one area where `tmap` is much better than `ggplot2` - working with *raster* data.

If you've never heard the term before, just think of a raster as image data. They're usually used to show roads & buildings on maps. `tmap` supports raster data out of the box, at the time of writing `ggplot2` doesn't & you need to do some dodgy work arounds to get raster data working. Plotting rasters in `tmap` is done using the `tm_rgb` function.

Here's a close up of Manchester CCG:

```{r warning=F, message=F}
shp_manc = shp_ccg %>%
filter(CCG19NM == 'NHS Manchester CCG')

tmap_mode('plot')
tm_shape(shp_manc) +
  tm_borders()
``` 

Unless you spend a lot of time looking at maps of Manchester, this probably isn't particularly useful. The easiest way to make this more useful would be to add in some of the major roads & landmarks, so the map shows more info about Manchester. Geographic information for more or less anywhere is collected through [OpenStreetMap](https://www.openstreetmap.org/). OpenStreetMap is a crowdsourced map of the world, anyone can make an account and add information about a road, building, park, or any other point of interest.

We can get rasters of OpenStreetMap using the `cartography` package. The function `getTiles` takes a shapefile & returns a raster of all OpenstreetMap data for that shapefile:

```{r warning=F, message=F}
osm_raster = getTiles(shp_manc, type = 'osm', 
                      zoom = 12, crop = T)                      
``` 

This next bit of code snips the raster to just the bit inside the shapefile. `mask` returns another raster, but any parts of the original raster which are outside of the shapefile are set to `NA`. We then `crop` the shapefile down to match the shapefile:

```{r warning=F, message=F}
msk = mask(osm_raster, shp_manc)
base_map = crop(msk, shp_manc)
```
Let's have a look. The left plot shows the full raster returned by the `getTiles` function, and the right plot shows the raster cropped to just Manchester CCG:

```{r warning=F, message=F}
full_map = tm_shape(osm_raster) +
  tm_rgb() +
  tm_shape(shp_manc) +
  tm_borders(lwd = 2) +
  tm_layout(frame = F)

cropped_map = tm_shape(base_map) +
  tm_rgb() +
  tm_layout(frame = F)

tmap_arrange(full_map, cropped_map)
```

Slightly more code than before, but there's nothing new here. Plotting rasters works the same way as plotting shapefiles - first say you want to draw a shape `tm_shape()`, then say how you want to display the shape `tm_rgb()`. The `tmap_arrange` function lets you arrange multiple maps into a single plot.

# Wrap up

This was a very quick overview of using `tmap`. If you want to learn more, here's a few resources I've found useful:

* The [`tmap` vignette](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)
* [Geocomputation with R](https://geocompr.robinlovelace.net/) by Lovelace, Nowosad, and Muenchow. This is a full on book about geocomputation in R. [Chapter 8](https://geocompr.robinlovelace.net/adv-map.html) goes into tmap in loads of detail